Testing Framework and Reference

### 2025-05-12 nbs/development/testing.qmd
docs(development): add integration tests
- note previously specified and unimplemented integration tests

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

```
src/pyrovelocity/tests/
├── __init__.py
├── conftest.py                # Shared pytest fixtures and configuration
├── fixtures/                  # Test data generation utilities
├── models/                    # Tests for model components
│   ├── __init__.py
│   ├── modular/               # Tests for modular implementation
│   │   ├── __init__.py
│   │   ├── conftest.py        # Modular-specific fixtures
│   │   ├── components/        # Unit tests for component implementations
│   │   │   ├── test_dynamics.py
│   │   │   ├── test_guides.py
│   │   │   ├── test_likelihoods.py
│   │   │   ├── test_observations.py
│   │   │   └── test_priors.py
│   │   ├── integration/       # BDD step implementations
│   │   │   ├── __init__.py
│   │   │   ├── conftest.py    # BDD-specific fixtures
│   │   │   ├── test_dynamics_model_steps.py
│   │   │   ├── test_model_steps.py
│   │   │   └── test_prior_model_steps.py
│   │   ├── selection/         # Tests for model selection
│   │   ├── utils/             # Tests for utility modules
│   │   ├── test_anndata_integration.py
│   │   ├── test_comparison.py
│   │   ├── test_factory.py
│   │   ├── test_interfaces.py
│   │   ├── test_model.py
│   │   ├── test_registry.py
│   │   └── test_unified_config.py
│   ├── jax/                   # Tests for JAX implementation (not yet documented)
│   └── test_*.py              # Tests for legacy implementation
├── features/                  # BDD feature specifications
│   └── models/
│       └── modular/
│           ├── dynamics_model.feature
│           ├── guide_model.feature
│           ├── likelihood_model.feature
│           ├── model.feature
│           ├── observation_model.feature
│           └── prior_model.feature
├── plots/                     # Tests for plotting functions
├── utils/                     # Tests for utility functions
├── validation/                # Tests for validation framework
└── test_*.py                  # Package-level tests
```

@@

| Module | Feature | Status | Description |
|--------|---------|--------|-------------|
| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts |
| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values |
| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation |
| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully |
| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction |
| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes |
| likelihood_model | Poisson likelihood model creates observation nodes | ✅ Implemented | Verify that the PoissonLikelihoodModel creates appropriate observation nodes |
| likelihood_model | Legacy likelihood model matches legacy implementation | ✅ Implemented | Verify that the LegacyLikelihoodModel matches the legacy implementation |
| observation_model | Standard observation model preprocesses data | ✅ Implemented | Verify that the StandardObservationModel correctly preprocesses data |
| guide_model | Auto guide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
| guide_model | Legacy auto guide factory matches legacy implementation | ✅ Implemented | Verify that the LegacyAutoGuideFactory matches the legacy implementation |
| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components |
| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects |

@@

### Example Feature File

@@

@@

Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory:

@@

>>>>

```
src/pyrovelocity/tests/
├── __init__.py
├── conftest.py                # Shared pytest fixtures and configuration
├── fixtures/                  # Test data generation utilities
├── models/                    # Tests for model components
│   ├── __init__.py
│   ├── modular/               # Tests for modular implementation
│   │   ├── __init__.py
│   │   ├── conftest.py        # Modular-specific fixtures
│   │   ├── components/        # Unit tests for component implementations
│   │   │   ├── test_dynamics.py
│   │   │   ├── test_guides.py
│   │   │   ├── test_likelihoods.py
│   │   │   ├── test_observations.py
│   │   │   └── test_priors.py
│   │   ├── integration/       # BDD step implementations
│   │   │   ├── __init__.py
│   │   │   ├── conftest.py    # BDD-specific fixtures
│   │   │   ├── test_component_integration.py
│   │   │   ├── test_dynamics_model_steps.py
│   │   │   ├── test_guide_model_steps.py
│   │   │   ├── test_likelihood_model_steps.py
│   │   │   ├── test_model_steps.py
│   │   │   ├── test_observation_model_steps.py
│   │   │   └── test_prior_model_steps.py
│   │   ├── selection/         # Tests for model selection
│   │   ├── utils/             # Tests for utility modules
│   │   ├── test_anndata_integration.py
│   │   ├── test_comparison.py
│   │   ├── test_factory.py
│   │   ├── test_interfaces.py
│   │   ├── test_model.py
│   │   ├── test_registry.py
│   │   └── test_unified_config.py
│   ├── jax/                   # Tests for JAX implementation (not yet documented)
│   └── test_*.py              # Tests for legacy implementation
├── features/                  # BDD feature specifications
│   └── models/
│       └── modular/
│           ├── dynamics_model.feature
│           ├── guide_model.feature
│           ├── likelihood_model.feature
│           ├── model.feature
│           ├── observation_model.feature
│           └── prior_model.feature
├── plots/                     # Tests for plotting functions
├── utils/                     # Tests for utility functions
├── validation/                # Tests for validation framework
└── test_*.py                  # Package-level tests
```

@@

| Module | Feature | Status | Description |
|--------|---------|--------|-------------|
| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts |
| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values |
| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation |
| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully |
| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction |
| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes |
| likelihood_model | Poisson likelihood model defines distributions | ✅ Implemented | Verify that the PoissonLikelihoodModel defines appropriate Poisson distributions |
| likelihood_model | Legacy likelihood model matches legacy implementation | ✅ Implemented | Verify that the LegacyLikelihoodModel matches the legacy implementation |
| likelihood_model | Likelihood model with scaling factors | ✅ Implemented | Verify that the likelihood model correctly incorporates scaling factors |
| likelihood_model | Likelihood model with zero counts | ✅ Implemented | Verify that the likelihood model handles zero counts gracefully |
| likelihood_model | Likelihood model with Pyro plates | ✅ Implemented | Verify that the likelihood model correctly uses Pyro plates for batch dimensions |
| observation_model | Standard observation model transforms data | ✅ Implemented | Verify that the StandardObservationModel correctly transforms input data |
| observation_model | Observation model computes size factors | ✅ Implemented | Verify that the observation model correctly computes size factors |
| observation_model | Observation model handles missing values | ✅ Implemented | Verify that the observation model handles missing values gracefully |
| observation_model | Observation model with normalization | ✅ Implemented | Verify that the observation model correctly normalizes data |
| observation_model | Observation model preserves data integrity | ✅ Implemented | Verify that the observation model preserves biological signal and doesn't introduce artifacts |
| guide_model | AutoGuide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
| guide_model | Legacy auto guide factory matches legacy implementation | ✅ Implemented | Verify that the LegacyAutoGuideFactory matches the legacy implementation |
| guide_model | Guide with parameter initialization | ✅ Implemented | Verify that the guide correctly initializes parameters with specified scale |
| guide_model | Guide with different variational families | ✅ Implemented | Verify that the guide supports different variational families |
| guide_model | Guide with custom initialization | ✅ Implemented | Verify that the guide supports custom initialization functions |
| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components |
| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects |

@@

### Example Feature Files

PyroVelocity uses feature files to describe the expected behavior of each component. Here are examples from different components:

#### Dynamics Model Example

@@

#### Inference Guide Example

```gherkin
Feature: Inference Guide
  As a computational biologist
  I want to define variational distributions for RNA velocity inference
  So that I can perform efficient approximate Bayesian inference

  Scenario: Guide with different variational families
    Given I have an AutoGuideFactory with guide_type="<guide_type>"
    When I create a guide for the model
    Then the guide should be an instance of <guide_class>
    And the guide should have the appropriate variational family

    Examples:
      | guide_type           | guide_class                  |
      | AutoNormal           | AutoNormal                   |
      | AutoDiagonalNormal   | AutoDiagonalNormal           |
      | AutoLowRankMultivariateNormal | AutoLowRankMultivariateNormal |
      | AutoDelta            | AutoDelta                    |
```

#### Likelihood Model Example

```gherkin
Feature: Likelihood Model
  As a computational biologist
  I want to define likelihood distributions for RNA velocity observations
  So that I can relate expected counts to observed counts in single-cell data

  Scenario: Poisson likelihood model defines distributions
    Given I have a PoissonLikelihoodModel
    When I run the forward method
    Then the model should define Poisson distributions for unspliced and spliced counts
    And the distributions should use the expected counts as rate parameters
    And the model should register observations with Pyro
```

#### Observation Model Example

```gherkin
Feature: Observation Model
  As a computational biologist
  I want to preprocess and transform RNA velocity data
  So that it can be used effectively in the model

  Scenario: Observation model with normalization
    Given I have a StandardObservationModel with normalization
    When I run the forward method
    Then the model should normalize the data
    And the normalized data should have the expected properties
    And the context should include the normalized data
```

@@

Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory. PyroVelocity has implemented step definitions for all the major components:

- `test_dynamics_model_steps.py`: Implements steps for testing dynamics models
- `test_prior_model_steps.py`: Implements steps for testing prior models
- `test_likelihood_model_steps.py`: Implements steps for testing likelihood models
- `test_observation_model_steps.py`: Implements steps for testing observation models
- `test_guide_model_steps.py`: Implements steps for testing inference guides
- `test_model_steps.py`: Implements steps for testing the full PyroVelocityModel

Additionally, there's a `test_component_integration.py` file that tests the interactions between different components to ensure they work correctly together.

Here's an example of a step definition file:

@@

### Component Integration Tests

In addition to the BDD step definitions, PyroVelocity includes component integration tests that verify the interactions between different components:

```python
# src/pyrovelocity/tests/models/modular/integration/test_component_integration.py
class TestComponentIntegration:
    """Tests for component interactions."""

    def test_dynamics_prior_integration(self):
        """Test that dynamics model and prior model work together."""
        # Create component models
        dynamics_model = StandardDynamicsModel()
        prior_model = LogNormalPriorModel()

        # Sample parameters from prior
        n_genes = 5
        prior_params = prior_model.sample_parameters(n_genes=n_genes)

        # Use parameters with dynamics model
        u0 = torch.zeros(n_genes)
        s0 = torch.zeros(n_genes)

        # Simulate dynamics
        times, u, s = dynamics_model.simulate(
            u0=u0,
            s0=s0,
            alpha=prior_params["alpha"],
            beta=prior_params["beta"],
            gamma=prior_params["gamma"],
            t_max=10.0,
            n_steps=100,
        )

    def test_full_model_integration(self, simple_data):
        """Test that all components work together in the full model."""
        # Create components
        dynamics_model = StandardDynamicsModel()
        prior_model = LogNormalPriorModel()
        likelihood_model = PoissonLikelihoodModel()
        observation_model = StandardObservationModel()
        guide_model = AutoGuideFactory(guide_type="AutoNormal")

        # Create the full model
        model = PyroVelocityModel(
            dynamics_model=dynamics_model,
            prior_model=prior_model,
            likelihood_model=likelihood_model,
            observation_model=observation_model,
            guide_model=guide_model,
        )

        # Call the model with test data
        result = model.forward(
            u_obs=simple_data["u_obs"],
            s_obs=simple_data["s_obs"]
        )
```
